# eCLIP analysis

## eCLIP mapping
Maps to a 'oneCopy' genome (with one copy of repeated snoRNAs) and a 'unique' genome (unique copies of repeated snoRNAs)

### Clone eclip_sno into PIPELINE repo
```{bash, eval=FALSE}
cd /n/whipple_lab/share_root/Users/cwhilden/PIPELINE
git clone git@github.com:whildencm/eclip_sno.git
```

### Prepare config files
-Copied test files from PIPELINE/eclip_sno into config/
-Copied slurm profile into profiles/

### Execute the pipeline
```{bash, eval=FALSE}
cd /n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2
module load Mambaforge
mamba activate /n/whipple_lab/share_root/Users/cwhilden/ENVS/snakemake_py311
pipeline=/n/whipple_lab/share_root/Users/cwhilden/PIPELINE/eclip_sno

# DRY RUN
snakemake --snakefile $pipeline/eclip.smk --configfile config/eclip_config.yaml --dryrun --unlock

# FULL RUN: SLURM
snakemake --profile profiles/slurm_eclip \
  --snakefile $pipeline/eclip.smk \
  --configfile config/eclip_config.yaml

mamba deactivate
```

## Peak calling
- Notes on how I generated the custom species files for clipper are in scripts/clipper_envBuild_mm39.Rmd

```{bash, eval=FALSE}
out=results/eCLIP/peakCalling/map_oneCopy
bamDir=results/eCLIP/map_oneCopy

sbatch scripts/eCLIP_peakCalling.sh $bamDir/SSEXPT8_IP_B_S4_R1_001.trim.Aligned.sortedByCoord.out.rmdup.bam $out
sbatch scripts/eCLIP_peakCalling.sh $bamDir/Fbl_IP_S2_L12_R1_001.trim.Aligned.sortedByCoord.out.rmdup.bam $out
sbatch scripts/eCLIP_peakCalling.sh $bamDir/SSEXPT8_IP_A_S3_R1_001.trim.Aligned.sortedByCoord.out.rmdup.bam $out
sbatch scripts/eCLIP_peakCalling.sh $bamDir/TDZ-7_S7_R1_001.trim.Aligned.sortedByCoord.out.rmdup.sorted.bam $out

```

### Merge peak coordinates of biological replicates
Intersect - keep the intersecting peak coordinates of RepA and RepB

```{bash, eval=FALSE}
salloc -p test -c 4 --mem 50G -t 300
module load Mambaforge
mamba activate /net/rcstorenfs/ifs/rc_labs/whipple_lab/share_root/Users/cwhilden/ENVS/chimericCW_001

peak=results/eCLIP/peakCalling/map_oneCopy

fileA=$peak/SSEXPT8_IP_A_S3_R1_001.bed
fileB=$peak/SSEXPT8_IP_B_S4_R1_001.bed

# Output "intersection" file
bedtools intersect -a "$fileA" -b "$fileB" \
  | bedtools sort -i - \
  | bedtools merge -d 1 -c 4,5,6,7,8 -o distinct,min,distinct,min,max \
   > "${peak}/iNeuron_rep_coord_intersect.bed"

# Merge adjacent peak coordinates
bedtools sort -i $peak/Fbl_IP_S2_L12_R1_001.bed | bedtools merge -d 1 -c 4,5,6,7,8 -o distinct,min,distinct,min,max > $peak/Fbl_IP_S2_L12_R1_001.merge.bed
bedtools sort -i $peak/TDZ-7_S7_R1_001.trim.Aligned.sortedByCoord.out.rmdup.sorted.bam.bed| bedtools merge -d 1 -c 4,5,6,7,8 -o distinct,min,distinct,min,max > $peak/TDZ-7_S7_R1_001.merge.bed



```
### Annotate peaks with gene info:

mamba activate clipper3_mm39

#using homer
gtf=/n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2/references/gencode.vM38.primary_assembly.annotation.gtf

cd results/eCLIP/peakCalling/map_oneCopy
annotatePeaks.pl iNeuron_rep_coord_intersect.bed none -gtf $gtf > "iNeuron_rep_coord.annotated.bed"
annotatePeaks.pl iNeuron_rep_coord_intersect.bed none -gtf $gtf > "iNeuron_rep_coord.annotated.bed"

annotatePeaks.pl Fbl_IP_S2_L12_R1_001.merge.bed none -gtf $gtf > "Fbl_IP_S2_L12_R1_001.merge.annotated.bed"
annotatePeaks.pl TDZ-7_S7_R1_001.merge.bed none -gtf $gtf > "TDZ-7_S7_R1_001.merge.annotated.bed"


### Normalize peaks to input

```{bash, eval=FALSE}
module load Mambaforge
mamba activate IDR_py38
bamDir=results/eCLIP/map_oneCopy

sbatch scripts/eCLIP_peakNormalization.sh $bamDir/SSEXPT8_IP_A_S3_R1_001.trim.Aligned.sortedByCoord.out.rmdup.sorted.bam \
$bamDir/SSEXPT8_IN_A_S1_R1_001.trim.Aligned.sortedByCoord.out.rmdup.sorted.bam \
results/eCLIP/peakCalling/map_oneCopy/iNeuron_rep_coord_intersect.bed \
results/eCLIP

sbatch scripts/eCLIP_peakNormalization.sh $bamDir/SSEXPT8_IP_B_S4_R1_001.trim.Aligned.sortedByCoord.out.rmdup.sorted.bam \
$bamDir/SSEXPT8_IN_B_S2_R1_001.trim.Aligned.sortedByCoord.out.rmdup.sorted.bam \
results/eCLIP/peakCalling/map_oneCopy/iNeuron_rep_coord_intersect.bed \
results/eCLIP

sbatch scripts/eCLIP_peakNormalization.sh $bamDir/Fbl_IP_S2_L12_R1_001.trim.Aligned.sortedByCoord.out.rmdup.sorted.bam \
$bamDir/Fbl_IN_S1_L12_R1_001.trim.Aligned.sortedByCoord.out.rmdup.sorted.bam \
results/eCLIP/peakCalling/map_oneCopy/Fbl_IP_S2_L12_R1_001.merge.bed \
results/eCLIP

sbatch scripts/eCLIP_peakNormalization.sh $bamDir/TDZ-7_S7_R1_001.trim.Aligned.sortedByCoord.out.rmdup.sorted.bam \
$bamDir/TDZ-3_S3_R1_001.trim.Aligned.sortedByCoord.out.rmdup.sorted.bam \
results/eCLIP/peakCalling/map_oneCopy/TDZ-7_S7_R1_001.merge.bed \
results/eCLIP


```
























# counting stuff
Counting:
I copied the snoRNA names from Lab Computing Resiources / snoRNA_mmu/snoRNA_MASTERv5.txt
Uploaded to UCSC genome table browser and exported results > references/snoRNAs_mm39.txt & saf

There are a few ways I could do the counting. I'm going to start with the unique copies,
and count everything that does not multimap or overlap multiple gene ids.

```{bash, eval=FALSE}
salloc -p test -c 10 --mem 50G -t 300
module load Mambaforge
mamba activate /n/whipple_lab/share_root/Users/cwhilden/ENVS/chimericCW_001

saf=references/eclip_sno/sno_mm39.saf
bam=results/eCLIP/map_unique/SSEXPT8_IN_A_S1_R1_001.trim.Aligned.sortedByCoord.out.bam
outDir=results/eCLIP/counts_unique
mkdir -p $outDir

featureCounts -T 8 -F SAF -a $saf -o $outDir/SSEXPT8_IN_A_S1_R1_001.counts.unique.txt $bam



```



### one copy counting
```{bash, eval=FALSE}
salloc -p test -c 10 --mem 50G -t 300
module load Mambaforge
mamba activate /n/whipple_lab/share_root/Users/cwhilden/ENVS/chimericCW_001

saf=references/eclip_sno/sno_mm39.saf
bamDir=results/eCLIP/map_oneCopy
outDir=results/eCLIP/counts_oneCopy
mkdir -p $outDir

for bam in $bamDir/*.out.rmdup.bam; do
  # no multi-mappers or multi-overlap

  base=$(basename "$bam" ".trim.Aligned.sortedByCoord.out.rmdup.sorted.bam")
  featureCounts -T 8 -F SAF -a $saf -o ${outDir}/${base}.counts.oneCopy.txt $bam

done

```

### settings
```{r}
library(tidyverse, quietly=TRUE)
library(ggplot2, quietly=TRUE)
library(tools, quietly=TRUE)
library(ggrepel, quietly=TRUE)
```

```{r setup}
# Project directory
clusterRoot <- "/n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2"
macRoot     <- "/Volumes/whipple_lab/Users/cwhilden/PROJECTS/THESIS/chapter2"

# Import whilden plot settings
plotSave <- function(plotVar, fileName, outSave=outDir, w=1.5, h=1.5){
  ggsave(
  filename = file.path(outSave,fileName),
  plot = plotVar,
  device = "pdf",
  width = w,
  height = h,
  units = "in"
  )
}

# PLOT THEME
paletteCW <- c(
  Pink_CW      = "#EC008C",
  Teal_CW      = "#008C8C",
  Charcoal_CW  = "#2E2E2E",
  LimeGreen_CW = "#9BC53D",
  SoftOrange_CW= "#F4A261",
  Purple_CW    = "#7B2CBF"
)

themeCW <- function(baseSize = 6) {
  theme_minimal(base_size = baseSize) %+replace%
    theme(
      # panel
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(fill = NA, colour = "black", size = 0.5),
      
      # axes lines/ticks/text
      axis.line = element_blank(),                      # border handles the visual frame
      axis.ticks = element_line(size = 0.5),
      axis.ticks.length = unit(0.03, "in"),
      axis.text = element_text(size = baseSize, colour = "black"),
      axis.title = element_text(size = baseSize, colour = "black"),
      
      # legend
      legend.key        = element_blank(),
      legend.background = element_blank(),
      legend.text       = element_text(size = baseSize),
      legend.title = element_blank(),
      
      legend.key.size   = unit(0.2, "cm"),
      legend.spacing.y  = unit(1, "pt"),
      legend.margin     = margin(0, 0, 0, 0, unit = "pt"),
      legend.position = "top",
      
      # plot text & margins
      plot.title = element_text(size = baseSize, colour = "black"),
      plot.subtitle = element_text(size = baseSize, colour = "black"),
      plot.margin = margin(5, 5, 5, 5, unit = "pt"),
      
      aspect.ratio=1
    )
  
}
```

### summary file
```{r}
# Define paths
sampleInfoFile<-file.path(macRoot,"references/sampleAlias.csv")
countDir<-file.path(macRoot,"results/eCLIP/counts_oneCopy")

###########
# Function to combine counts across multiple samples into one data frame
# Read in feature counts data
dataFiles <- list.files(path = countDir, 
                        pattern = "\\.counts.oneCopy.txt$", 
                        full.names = TRUE)


# Extract sample/snoRNA information from filename
names<-sub("\\.trim.Aligned.sortedByCoord.out.rmdup.bam.counts.oneCopy.txt","",basename(dataFiles))

# generate a named list of counts
data <- purrr::set_names(dataFiles, names) %>%
  purrr::map(~ readr::read_tsv(.x, show_col_types = FALSE, skip=1))
  
# Merge named list into a single data frame
allData <- data[[1]][,1, drop=FALSE]

for (i in seq_along(names)) {
  # Read in counts column (column 7)
  allData[[names[i]]] <- data[[i]][,7, drop=TRUE]
  
} 

outDir<-file.path(macRoot,"summaryFiles/eCLIP")
dir.create(outDir, recursive = TRUE)

write.csv(allData, file.path(outDir,"eCLIP_oneCopy_RAW_CountsSummary.csv"))


# Get CPM
cpm_data <- allData %>%
  mutate(across(-Geneid, ~ . / sum(.) * 1e6))

write.csv(cpm_data, file.path(outDir,"eCLIP_oneCopy_CPM_CountsSummary.csv"))
```




