---
title: "Nm_snoRNAs_chimericAnalyses.Rmd"
author: "Courtney Whilden"
date: "2026-03-11"
output: html_document
---
# Analysis of Nm-snoRNA chimeric reads
## Overview

### Set up
- Loads packages
```{r loadPackages}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse, quietly=TRUE)
library(ggplot2, quietly=TRUE)
library(metagene2, quietly=TRUE)
library(ggrepel, quietly=TRUE)

```

- Sets project root directories
- Sets plot themes
```{r setup}
# Project directory
clusterRoot <- "/n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2"
macRoot     <- "/Volumes/whipple_lab/Users/cwhilden/PROJECTS/THESIS/chapter2"

# Import whilden plot settings
plotSave <- function(plotVar, fileName, outSave=outDir, w=1.5, h=1.5){
  ggsave(
  filename = file.path(outSave,fileName),
  plot = plotVar,
  device = "pdf",
  width = w,
  height = h,
  units = "in"
  )
}

# PLOT THEME
paletteCW <- c(
  Pink_CW      = "#EC008C",
  Teal_CW      = "#008C8C",
  Charcoal_CW  = "#2E2E2E",
  LimeGreen_CW = "#9BC53D",
  SoftOrange_CW= "#F4A261",
  Purple_CW    = "#7B2CBF"
)

themeCW <- function(baseSize = 6) {
  theme_minimal(base_size = baseSize) %+replace%
    theme(
      # panel
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(fill = NA, colour = "black", size = 0.5),
      
      # axes lines/ticks/text
      axis.line = element_blank(),                      # border handles the visual frame
      axis.ticks = element_line(size = 0.5),
      axis.ticks.length = unit(0.03, "in"),
      axis.text = element_text(size = baseSize, colour = "black"),
      axis.title = element_text(size = baseSize, colour = "black"),
      
      # legend
      legend.key        = element_blank(),
      legend.background = element_blank(),
      legend.text       = element_text(size = baseSize),
      legend.title = element_blank(),
      
      legend.key.size   = unit(0.2, "cm"),
      legend.spacing.y  = unit(1, "pt"),
      legend.margin     = margin(0, 0, 0, 0, unit = "pt"),
      legend.position = "top",
      
      # plot text & margins
      plot.title = element_text(size = baseSize, colour = "black"),
      plot.subtitle = element_text(size = baseSize, colour = "black"),
      plot.margin = margin(5, 5, 5, 5, unit = "pt"),
      
      aspect.ratio=1
    )
  
}
```

# Calculate chimeras per target class - Nm sno vs. 116

Summary file paths
```{r}
# Read in Nm snoRNA annotation file
Nm <- read_csv(file.path(macRoot,"references/Known_Nm_snoRNAs.csv"))%>%
  select('SNO')

# import chimeras, and separate into classes "Nm", "Snord116", "other"
chimeras <- read_csv(file.path(macRoot,"summaryFiles/chimericCounts/All_ChimeraCounts.csv")) %>%
  # strip ".sno" suffix from sample columns if present
  rename_with(~ sub("\\.sno$", "", .x), -c(snoRNA, target)) %>%
  mutate(snoType=case_when(snoRNA %in% Nm$SNO ~ 'Nm',
                   snoRNA == 'Snord116_Snord116l2' ~ 'Snord116',
                   TRUE ~ 'Other'))

#Import sample alias file
alias <- read_csv(file.path(macRoot,"references/sampleAlias.csv"))

# Pivot for plotting & replace sample
chimeras_long <- chimeras %>%
  pivot_longer(
    cols = -c(snoRNA, target, snoType),
    names_to = "sample",
    values_to = "count"
  ) %>%
  left_join(alias, by = "sample") %>%
  mutate(alias_sno=paste(alias,snoType,sep=";")) %>%
  select(all_of(c('alias_sno','sample','alias','snoRNA','target','count')))

# Total chimeric reads per sample (CPM denominator)
sample_totals_bySnoClass <- chimeras_long %>%
  group_by(alias_sno) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop")

# Sum across all snoRNAs for each target/sample, then calculate CPM
chimeras_by_target <- chimeras_long %>%
  group_by(alias_sno,target) %>%
  summarise(target_count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  left_join(sample_totals_bySnoClass , by = "alias_sno") %>%
  mutate(cpm = (target_count / total_count) * 1e6) %>%
  separate(alias_sno, into = c("alias", "snoType"), sep = ";")


# Save raw counts per target for downstream analyses
rawPerTarget <- chimeras_long %>%
  group_by(sample) %>%
  summarise(total=sum(count,na.rm=TRUE),.groups="drop")

```

# Nm sno only
Exports a plot of the number of chimeras within each target class
```{r}
samples1 <- c('ESC_IP_RepA','iNeuron_IP_RepA','iNeuron_IP_RepB','Steric_noASO_RepA')

# Select samples & Nm snoRNAs only
Nm_plot <- chimeras_by_target %>%
  filter(alias %in% samples1 & snoType == "Nm") %>%
  mutate(
    alias   = factor(alias,   levels = samples1),
    target  = factor(target,  levels = c("genome", "scaRNA", "snRNA", "tRNA","rRNA"))
  )

# Plot
plot1<-ggplot(Nm_plot, aes(x = alias, y = cpm, fill = target)) +
  geom_bar(stat = "identity") +
  labs(
    y = "Read counts",
    x = "Sample",
    fill = "Category"
  ) +
  themeCW()

write.csv(Nm_plot,file.path(macRoot,"summaryFiles/chimericCounts/Nm_snoRNAs_chimera_target_distribution.csv"))
outDir=file.path(macRoot,"plots/Nm-snoRNAs")
if (!dir.exists(outDir)) dir.create(outDir, recursive = TRUE)

outDir <- file.path(macRoot,"plots/Nm-snoRNAs")
dir.create(outDir, recursive = TRUE)

plotSave(plot1, "Nm_snoRNAs_chimera_target_distribution.pdf")

plot1
```

#=======
# CHIMERAS PER SNORNA
#=======
```{r}
# Total chimeric reads per sample (CPM denominator)
sample_totals <- chimeras_long %>%
  group_by(alias) %>%
  summarise(total_count = sum(count, na.rm = TRUE), .groups = "drop")

# Sum across all snoRNAs for each target/sample, then calculate CPM
chimeras_by_sno <- chimeras_long %>%
  group_by(alias,snoRNA) %>%
  summarise(target_count = sum(count, na.rm = TRUE), .groups = "drop") %>%
  left_join(sample_totals, by = "alias") %>%
  mutate(cpm = (target_count / total_count) * 1e6) %>%
  mutate(log2CPMplus1=log2(cpm+1))

# Pivot wider so each alias is a column
chimeras_by_sno_wide <- chimeras_by_sno %>%
  select(alias, snoRNA, log2CPMplus1) %>%
  pivot_wider(names_from = alias, values_from = log2CPMplus1) %>%
  mutate(iNeuron_IP_Avg=(iNeuron_IP_RepA + iNeuron_IP_RepB)/2,
         sno_class=case_when(snoRNA %in% Nm$SNO ~ "Nm",
                             snoRNA == 'Snord116_Snord116l2' ~ 'Snord116',
                             TRUE ~ 'Other'))

snord116_CPM <- chimeras_by_sno %>%
  select(alias, snoRNA, cpm) %>%
  pivot_wider(names_from = alias, values_from = cpm) %>%
  mutate(iNeuron_IP_Avg=(iNeuron_IP_RepA + iNeuron_IP_RepB)/2) %>%
  filter(snoRNA == 'Snord116_Snord116l2')
      

# iN reps
# Calculate Pearson's r on the non-NA pairs
cor_val1 <- cor(
  chimeras_by_sno_wide$iNeuron_IP_RepA,
  chimeras_by_sno_wide$iNeuron_IP_RepB,
  use = "complete.obs",
  method = "pearson"
)

plot2 <- ggplot(chimeras_by_sno_wide, aes(x = iNeuron_IP_RepA, y = iNeuron_IP_RepB, color=sno_class)) +
  geom_point(size = 1.5, alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50", size = 0.3) +
  annotate("text",
           x = Inf, y = -Inf,
           hjust = 1.1, vjust = -0.5,
           label = paste0("r = ", round(cor_val1, 3)),
           size = 6 / .pt,
           colour = "black") +
  labs(title = "log2(Chimeric RPM + 1)") +
  themeCW() +
  expand_limits(x = 0, y = 0)

#plotSave(plot2, "iNeuron_chimerasPerSno_RepA_vs_RepB.pdf")
plot2
```
### plot snoRNA chimeras in ESCs vs. iNs
```{r}
top5_iN <- chimeras_by_sno_wide %>%
  slice_max(iNeuron_IP_Avg, n = 5)

plot3 <- ggplot(chimeras_by_sno_wide, aes(x = ESC_IP_RepA, y = iNeuron_IP_Avg, color=sno_class)) +
  geom_point(size=1.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey50", size = 0.3) +
  geom_text_repel(data = top5_iN,
                  aes(label = snoRNA),
                  color = "black",
                  size = 1,
                  box.padding = 0.5,
                  max.overlaps = Inf) +
  labs(title = "log2(Chimeric RPM + 1)") +
  themeCW() +
  expand_limits(x = 0, y = 0)

plotSave(plot3,"chimerasPerSno_ESC_v_iNeuron.pdf")
plot3
```





#======
# Merge Nm-snoRNA bam files into one
#======
# Find all bam files for a given sample and export a list
```{r, echo=FALSE}
# Project directory
clusterRoot <- "/n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2"
macRoot     <- "/Volumes/whipple_lab/Users/cwhilden/PROJECTS/THESIS/chapter2"

# Read in data files
Nm <- read_csv(file.path(macRoot,"references/Known_Nm_snoRNAs.csv"))%>%
  select('SNO')
  
bamDir <- file.path(macRoot,"results/mapChimeras")
  

output=file.path(macRoot,"results/Nm_snoRNAs/bamLists")
dir.create(output, recursive = TRUE)

# Generate lists of Nm-sno bam files for each sample
listNmBams <- function(sample,
                       bams=bamDir,
                       names=Nm,
                       outDir=output,
                       cluster=file.path(clusterRoot,"results/mapChimeras")){

  
  bamFiles=list.files(bams,
                      pattern=paste0("^", sample, "\\.snoRNA\\..+\\.chimeras\\.bam$"),
                      full.names=TRUE)
  

  # Extract target type from bam file name
  base <- basename(bamFiles)
  
  # Append cluster dir to path
  path=file.path(cluster,base)
  
  targetType <- sub(".*\\.snoRNA\\.([^.]+)\\..*\\.chimeras\\.bam$", "\\1", base)
  
  # Extract snoRNA name from bam file name
  snoName <- base
  snoName <- sub(".*\\.snoRNA\\.[^.]+\\.", "", snoName)
  snoName <- sub("\\.chimeras\\.bam$", "", snoName)
  
  # Combine into a data frame
  df_nm <- dplyr::filter(df, snoName %in% names$SNO)

  # Keep only Nm snoRNAs
  dfNm <- df[df$snoName %in% names$SNO, , drop = FALSE]
  
  # Write a Nm-snoRNA bam list for each target type
  targets <- sort(unique(dfNm$targetType))
  
  for (t in targets) {
    paths <- dfNm$path[dfNm$targetType == t]
    # stable ordering
    paths <- sort(paths)
    outPath <- file.path(outDir, paste0(sample,"_NmSno_", t, ".txt"))
    writeLines(paths, outPath)
    message(sprintf("Wrote %s (%d files)", outPath, length(paths)))
  }

}

listNmBams(sample="SSEXPT8_IP_A_S3_R1_001")
listNmBams(sample="SSEXPT8_IP_B_S4_R1_001")

```

## Merge Nm-sno bam files for each sample, based on bamLists
```{bash, eval=FALSE}
salloc -p whipple -c 10 --mem 50G -t 300

# Load environment
module load Mambaforge
mamba activate /n/whipple_lab/share_root/Users/cwhilden/ENVS/chimericCW_001     

# Set directories
clusterRoot=/n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2
listDir="${clusterRoot}/results/Nm_snoRNAs/bamLists"
outDir="${clusterRoot}/results/Nm_snoRNAs"
mkdir -p $outDir

for listFile in "$listDir"/*_NmSno_*.txt; do
  base="$(basename "$listFile")"

  # Parse sample and targetType from: <sample>_NmSno_<target>.txt
  sample="${base%_NmSno_*}"
  targetPart="${base#*_NmSno_}"
  targetType="${targetPart%.txt}"

  outBam="${outDir}/${sample}.All-NmSno-chimeras.${targetType}.bam"

  # Read paths from list (safe for spaces)
  mapfile -t bamFiles < "$listFile"

  if [[ ${#bamFiles[@]} -eq 0 ]]; then
    echo "Empty list: $listFile (skipping)"
    continue
  fi
  
  echo "Merging ${#bamFiles[@]} BAMs for sample=${sample} target=${targetType}"
  echo "  list: $listFile"
  echo "  out : $outBam"

  samtools merge -@ 8 -f "$outBam" "${bamFiles[@]}"
  samtools index -@ 8 "$outBam"
done

echo "Done. Merged BAMs in: $outDir"
```

## Generate chimeric RPM BW files for IGV tracks
```{bash, eval=FALSE}
# Start interactive session
salloc -p whipple -c 4 --mem 50G -t 300

# Load environment
module load Mambaforge
mamba activate /n/whipple_lab/share_root/Users/cwhilden/ENVS/chimericCW_001 

# Run chimeric bam2bw script on desired bam file:
totalChimerasFile="summaryFiles/chimericCounts/ChimerasPerClass_AllSamples.csv"
bamDir="results/Nm_snoRNAs"
bwDir="results/Nm_snoRNAs/CPM_BW"
mkdir -p $bwDir

# make rRNA and snRNA tracks
samples=(SSEXPT8_IP_A_S3_R1_001 SSEXPT8_IP_B_S4_R1_001)

for sample in "${samples[@]}"; do

  # rRNA chimeras
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.All-NmSno-chimeras.rRNA.bam \
  ${bwDir}/${sample}.All-NmSno-chimeras.rRNA.cpm.bw \
  ${bwDir}/${sample}.All-NmSno-chimeras.rRNA.cpm.log

  #snRNA chimeras
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.All-NmSno-chimeras.snRNA.bam \
  ${bwDir}/${sample}.All-NmSno-chimeras.snRNA.cpm.bw \
  ${bwDir}/${sample}.All-NmSno-chimeras.snRNA.cpm.log  

done
```

#==========
# METAGENE PLOTS : CHIMERAS NEAR NM SITES
#===========
### rRNA metagene plots
```{r}
nm_bed<-file.path(macRoot,'references/ModSites_rRNA_Nm.bed')
psu_bed<-file.path(macRoot,'references/ModSites_rRNA_pU.bed')

chimeraDir=file.path(macRoot,'results/Nm_snoRNAs')

bams <- c(
  file.path(chimeraDir, "SSEXPT8_IP_A_S3_R1_001.All-NmSno-chimeras.rRNA.bam"),
  file.path(chimeraDir, "SSEXPT8_IP_B_S4_R1_001.All-NmSno-chimeras.rRNA.bam")
)

makeMetagene <- function(bed,
                         chimeraBams){
  
  # Compare repA to repB
  design <- data.frame(
  Samples = chimeraBams,
  RepA = c(1, 0),
  RepB = c(0, 1)
  )
  
  # Create metagene plot
  metagene <- metagene2$new(
  regions = bed,
  bam_files = chimeraBams,
  assay = "chipseq",
  padding_size = 25
  )
  
  metagene$group_coverages(design = design, normalization = 'RPM')
  metagene$bin_coverages(bin_count = 25)
  metagene$plot(title = 'Nm-rRNA', 
                region_names = NULL, 
                design_names = NULL, 
                x_label = NULL) + 
    themeCW() + 
    scale_y_continuous(limits = c(2200,10000))

}


outDir <- file.path(macRoot,"plots/Nm-snoRNAs")


makeMetagene(psu_bed,bams) %>%
  plotSave("Metagene_rRNA_pseudoU.pdf",outDir)

makeMetagene(nm_bed,bams) %>%
  plotSave("Metagene_rRNA_Nm.pdf",outDir)

```


### snRNA metagene plots
```{r}
nm_sn<-file.path(macRoot,'references/ModSites_snRNA_Nm.bed')
psu_sn<-file.path(macRoot,'references/ModSites_snRNA_pU.bed')

chimeraDir=file.path(macRoot,'results/Nm_snoRNAs')

bams <- c(
  file.path(chimeraDir, "SSEXPT8_IP_A_S3_R1_001.All-NmSno-chimeras.snRNA.bam"),
  file.path(chimeraDir, "SSEXPT8_IP_B_S4_R1_001.All-NmSno-chimeras.snRNA.bam")
)

makeMetageneSN <- function(bed,
                         chimeraBams){
  
  # Compare repA to repB
  design <- data.frame(
  Samples = chimeraBams,
  RepA = c(1, 0),
  RepB = c(0, 1)
  )
  
  # Create metagene plot
  metagene <- metagene2$new(
  regions = bed,
  bam_files = chimeraBams,
  assay = "chipseq",
  padding_size = 15
  )
  
  metagene$group_coverages(design = design, normalization = 'RPM')
  metagene$bin_coverages(bin_count = 15)
  metagene$plot(title = 'Nm-rRNA', 
                region_names = NULL, 
                design_names = NULL, 
                x_label = NULL) + 
    themeCW() + 
    scale_y_continuous(limits = c(15000,150000))

}


outDir <- file.path(macRoot,"plots/Nm-snoRNAs")


makeMetageneSN(psu_sn,bams) %>%
  plotSave("Metagene_snRNA_pseudoU.pdf",outDir)

makeMetageneSN(nm_sn,bams) %>%
  plotSave("Metagene_snRNA_Nm.pdf",outDir)

```













