---
title: "Nm_snoRNAs_peakCalling_Analyses.Rmd"
author: "Courtney Whilden"
date: "2026-03-12"
output: html_document
---
```{r}
# Project directory
clusterRoot <- "/n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2"
macRoot     <- "/Volumes/whipple_lab/Users/cwhilden/PROJECTS/THESIS/chapter2"

library(tidyverse, quietly=TRUE)
library(ggplot2, quietly=TRUE)
library(tools, quietly=TRUE)
library(ggrepel, quietly=TRUE)
library(pROC, quietly=TRUE)


# PLOT THEME
paletteCW <- c(
  Pink_CW      = "#EC008C",
  Teal_CW      = "#008C8C",
  Charcoal_CW  = "#2E2E2E",
  LimeGreen_CW = "#9BC53D",
  SoftOrange_CW= "#F4A261",
  Purple_CW    = "#7B2CBF"
)

themeCW <- function(baseSize = 6) {
  theme_minimal(base_size = baseSize) %+replace%
    theme(
      # panel
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(fill = NA, colour = "black", size = 0.5),
      
      # axes lines/ticks/text
      axis.line = element_blank(),                      # border handles the visual frame
      axis.ticks = element_line(size = 0.5),
      axis.ticks.length = unit(0.03, "in"),
      axis.text = element_text(size = baseSize, colour = "black"),
      axis.title = element_text(size = baseSize, colour = "black"),
      
      # legend
      legend.key        = element_blank(),
      legend.background = element_blank(),
      legend.text       = element_text(size = baseSize),
      legend.title = element_blank(),
      
      legend.key.size   = unit(0.2, "cm"),
      legend.spacing.y  = unit(1, "pt"),
      legend.margin     = margin(0, 0, 0, 0, unit = "pt"),
      legend.position = "top",
      
      # plot text & margins
      plot.title = element_text(size = baseSize, colour = "black"),
      plot.subtitle = element_text(size = baseSize, colour = "black"),
      plot.margin = margin(5, 5, 5, 5, unit = "pt"),
      
      aspect.ratio=1
    )
}


######################################
# Function to save plots in my preferred size
plotSave <- function(plotVar, fileName, outSave=outDir){
  ggsave(
  filename = file.path(outSave,fileName),
  plot = plotVar,
  device = "pdf",
  width = 1.5,
  height = 1.5,
  units = "in"
  )
}
```

## PEAK CALLING Nm-snoRNAs
Input data: summaryFiles/peakSummaryFiles

### Plot the fraction of chimeras in peak between replicates and perform ROC

```{r}
# Data import
data_repA <- read_csv(file.path(macRoot,"summaryFiles/peakSummaryFiles/SSEXPT8_IP_A_S3_R1_001_Significant_Peak_Results.csv"))
data_repB <- read_csv(file.path(macRoot,"summaryFiles/peakSummaryFiles/SSEXPT8_IP_B_S4_R1_001_Significant_Peak_Results.csv"))
# Read in Nm snoRNA annotation file

Nm <- read_csv(file.path(macRoot,"references/Known_Nm_snoRNAs.csv"))


```


```{r}
# Add peak ID containing snoRNA name and peak coordinates
addPeakID <- function(data) {
  
  data$peakID <- paste0(data$snoRNA,";",data$chromosome,"_",data$start,"_",data$end)
  
  return(data)
  
}

data_repA <- addPeakID(data_repA)
data_repB <- addPeakID(data_repB)

# Filter to peakID and fraction values
filterData <- function(data) {
  
  data2 <- data %>%
    select(all_of(c("peakID","fraction","log2_fold_change")))
  
  return(data2) 
  
}

A_filter <- filterData(data_repA)
colnames(A_filter)[2]<-"fracRepA"
colnames(A_filter)[3]<-"l2fcA"

B_filter <- filterData(data_repB)
colnames(B_filter)[2]<-"fracRepB"
colnames(B_filter)[3]<-"l2fcB"

# Join replicates into one data frame
data_join <- left_join(A_filter, B_filter, by='peakID')

data_join[is.na(data_join)] <- 0

# Annotate peaks as known or unknown
peakStatus <- select(data_repA,(all_of(c("peakID","peakStatus"))))

data_join <- left_join(data_join,peakStatus,by="peakID") %>% 
  mutate(avg_l2fc=(l2fcA+l2fcB)/2,
         avg_frac=(fracRepA+fracRepB)/2,
         # Add Nm-snoRNA annotations to pick out true positives and false positives
         is_nm = ifelse(sub(";.*", "", peakID) %in% Nm$SNO, "Nm", "not")
         ) 
```

### Fraction in peak, replicates
```{r}
cor_val_frac <- cor(
  data_join$fracRepA,
  data_join$fracRepB,
  use = "complete.obs",
  method = "pearson"
)

# Plot 
plot<-ggplot(data_join, aes(x = fracRepA, y = fracRepB, color = peakStatus)) +
  geom_point(alpha = 0.8, size = 1.5) +
  annotate("text",
           x = Inf, y = -Inf,
           hjust = 1.1, vjust = -0.5,
           label = paste0("r = ", round(cor_val_frac, 3)),
           size = 6 / .pt,
           colour = "black") +
  
  scale_color_manual(
    values = c(
      known = "#008C8C",
      other = "#2E2E2E"
    )
  ) +
  
  labs(
    x = "Fraction of chimeras in peak (Rep A)",
    y = "Fraction of chimeras in peak (Rep B)"
  ) +
  geom_vline(xintercept=0.1) +
  geom_hline(yintercept=0.1) +
  themeCW()

outDir=file.path(macRoot,"plots/Nm-snoRNAs")

plotSave(plot,"FractionChimeras_RepA_v_RepB.pdf")

```

### ROC to check accuracy of fraction threshold
True positives = Nm-snoRNA peaks overlapping with a known Nm-snoRNA interaction
False = Nm-snoRNA peaks NOT overlapping a known Nm-snoRNA interaction
  Overlapping = Nm site is within the start and end coordinates of the peak
  
```{r}
data_Nm <- data_join %>%
  filter(is_nm=="Nm")

rocAnalysis <- roc(
  response  = data_join$peakStatus,
  predictor = data_join$avg_frac,
  levels    = c("unknown", "known"),
  direction = "<"
)

# AUC
aucValue <- as.numeric(auc(rocAnalysis))

# Point at chosen threshold (0.1)
pt01 <- coords(
  rocAnalysis,
  x = 0.1,
  input = "threshold",
  ret = c("threshold", "specificity", "sensitivity"),
  transpose = FALSE
) %>% as.data.frame()

# Youden-best point
ptBest <- coords(
  rocAnalysis,
  x = "best",
  best.method = "youden",
  ret = c("threshold", "specificity", "sensitivity"),
  transpose = FALSE
) %>% as.data.frame()

# ROC plot data (specificity / sensitivity along thresholds)
rocDf <- coords(
  rocAnalysis,
  x = "all",
  ret = c("specificity", "sensitivity", "threshold"),
  transpose = FALSE
) %>% as.data.frame() %>%
  mutate(fpr = 1 - specificity)

# Plot
pRoc <- ggplot(rocDf, aes(x = fpr, y = sensitivity)) +
  geom_path(linewidth = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  # threshold = 0.1 point
  geom_point(
    data = pt01 %>% mutate(fpr = 1 - specificity),
    aes(x = fpr, y = sensitivity),
    size = 2.2
  ) +
  geom_text(
    data = pt01 %>% mutate(fpr = 1 - specificity),
    aes(x = fpr, y = sensitivity,
        label = paste0("t=0.1\nSe=", round(sensitivity, 2),
                       " Sp=", round(specificity, 2))),
    hjust = -0.05, vjust = -0.4, size = 3
  ) +
  # best (Youden) point
  geom_point(
    data = ptBest %>% mutate(fpr = 1 - specificity),
    aes(x = fpr, y = sensitivity),
    shape = 17, size = 2.4
  ) +
  geom_text(
    data = ptBest %>% mutate(fpr = 1 - specificity),
    aes(x = fpr, y = sensitivity,
        label = paste0("Youden best\n t=", signif(threshold, 3))),
    hjust = -0.05, vjust = 1.2, size = 3
  ) +
  coord_equal() +
  labs(
    x = "False positive rate (1 - specificity)",
    y = "True positive rate (sensitivity)",
    title = paste0("ROC: fraction in peak (AUC = ", round(aucValue, 3), ")")
  ) +
  themeCW()

plotSave(pRoc, "ROC_Nm_sno_fraction_in_peak.pdf")

pRoc

```


#### Plot fraction of chimeras vs. log2Peak Enrichment

```{r}
plot2<-ggplot(data_join, aes(x = avg_l2fc, y = avg_frac, color = peakStatus)) +
  geom_point(alpha = 0.8, size = 1.5) +

  scale_color_manual(
    values = c(
      known = "#008C8C",
      other = "#2E2E2E"
    )
  ) +
  
  labs(
    x = "log2( Peak fold enrichment) RepA",
    y = "Fraction of chimeras in peak (Rep A)"
  ) +
  geom_vline(xintercept=1.5) +
  geom_hline(yintercept=0.1) +
  themeCW()



plotSave(plot2,"FractionInPeak_v_peakEnrichment.pdf")

plot2
```

### Quantify significant known interactions

```{r}
getSigPeaks <- function(data){
  
  data2 <- data %>%
    filter(peakStatus=='known') %>%
    filter(fraction>=0.1 & log2_fold_change>=1.5) %>%
    select(all_of(c('peakID','targetClass')))

}

sigA <- getSigPeaks(data_repA)
sigB <- getSigPeaks(data_repB)
sigBoth <- intersect(sigA$peakID,sigB$peakID)

sigBothData <- sigA %>%
  filter(peakID %in% sigBoth)

# Calculate total number of significant peaks
sigPeakCount <- data.frame(TotalNm=nrow(Nm)) 

Nm_byClass <- data.frame(
  Nm_rRNA = sum(Nm$CLASS == 'rRNA'),
  Nm_snRNA = sum(Nm$CLASS == 'snRNA'),
  Nm_tRNA = sum(Nm$CLASS == 'tRNA'),
  Sig_rRNA = sum(sigBothData$targetClass == 'rRNA'),
  Sig_snRNA = sum(sigBothData$targetClass == 'snRNA'),
  Sig_tRNA = sum(sigBothData$targetClass == 'tRNA')
)

df <- Nm_byClass[1, ]

longDf <- df %>%
  pivot_longer(
    cols = everything(),
    names_to = c("type", "class"),
    names_sep = "_"
  ) %>%
  pivot_wider(names_from = type, values_from = value) %>%
  mutate(NotSig = Nm - Sig) %>%
  select(class, Sig, NotSig) %>%
  pivot_longer(
    cols = c(Sig, NotSig),
    names_to = "Significance",
    values_to = "Count"
  )

plot5 <- ggplot(longDf, aes(x = class, y = Count, fill = Significance)) +
  geom_bar(stat = "identity") +
  geom_text(
    aes(label = Count),
    position = position_stack(vjust = 0.5),  # centres label in each stack segment
    size = 2,
    colour = "white"
  ) +
  labs(
    x = "Target Class",
    y = "Count",
    fill = ""
  ) +
  themeCW()



plotSave(plot5, "Significant_Nm_interactions_counts.pdf")

plot5
```



