---
title: "Snord116_complementarity_analysis.Rmd"
author: "Courtney Whilden"
date: "2026-03-19"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse, quietly=TRUE)
library(ggrepel, quietly=TRUE)
library(ggplot2, quietly=TRUE)
library(pROC, quietly=TRUE)
# Project directory
clusterRoot <- "/n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2"
macRoot     <- "/Volumes/whipple_lab/Users/cwhilden/PROJECTS/THESIS/chapter2"

# Import whilden plot settings
plotSave <- function(plotVar, fileName, outSave=outDir, w=1.5, h=1.5){
  ggsave(
  filename = file.path(outSave,fileName),
  plot = plotVar,
  device = "pdf",
  width = w,
  height = h,
  units = "in"
  )
}

# PLOT THEME
paletteCW <- c(
  Pink_CW      = "#EC008C",
  Teal_CW      = "#008C8C",
  Charcoal_CW  = "#2E2E2E",
  LimeGreen_CW = "#9BC53D",
  SoftOrange_CW= "#F4A261",
  Purple_CW    = "#7B2CBF"
)

themeCW <- function(baseSize = 6) {
  theme_minimal(base_size = baseSize) %+replace%
    theme(
      # panel
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(fill = NA, colour = "black", size = 0.5),
      
      # axes lines/ticks/text
      axis.line = element_blank(),                      # border handles the visual frame
      axis.ticks = element_line(size = 0.5),
      axis.ticks.length = unit(0.03, "in"),
      axis.text = element_text(size = baseSize, colour = "black"),
      axis.title = element_text(size = baseSize, colour = "black"),
      
      # legend
      legend.key        = element_blank(),
      legend.background = element_blank(),
      legend.text       = element_text(size = baseSize),
      legend.title = element_blank(),
      
      legend.key.size   = unit(0.2, "cm"),
      legend.spacing.y  = unit(1, "pt"),
      legend.margin     = margin(0, 0, 0, 0, unit = "pt"),
      legend.position = "top",
      
      # plot text & margins
      plot.title = element_text(size = baseSize, colour = "black"),
      plot.subtitle = element_text(size = baseSize, colour = "black"),
      plot.margin = margin(5, 5, 5, 5, unit = "pt"),
      
      aspect.ratio=1
    )
}
```

## Get center 20 nt of each peak
```{r}
makeBedFile <- function(peaks, outFile, windowSize = 20) {
  
  # Middle of peak range set in function
    halfWindow <- windowSize / 2
  
    # If the peak width is <20, keep the whole peak
    # If peak width is >20, find the midpoint and extract +10 to -10
    peaks2 <- peaks %>%
        mutate(
            peakID = paste(chromosome,
                           start,
                           end,
                           sep = "_"),
            
            peakWidth = end - start,
            midPoint  = start + (peakWidth / 2),

            newStart = if_else(
                peakWidth >= windowSize,
                round(midPoint - halfWindow),
                start
            ),

            newEnd = if_else(
                peakWidth >= windowSize,
                round(midPoint + halfWindow),
                end
            ),

            # prevent negative coordinates
            newStart = pmax(0, newStart)
        )

    bed <- peaks2 %>%
        transmute(
            chrom = chromosome,
            start = as.integer(newStart),
            end   = as.integer(newEnd),
            name  = peakID
        )

    write.table(
        bed,
        file = outFile,
        sep = "\t",
        quote = FALSE,
        row.names = FALSE,
        col.names = FALSE
    )
}

peaks <- read_csv("/Volumes/whipple_lab/Users/cwhilden/PROJECTS/THESIS/chapter2/summaryFiles/peakSummaryFiles/TDZ-7_S7_R1_001_Significant_Peak_Results.csv", show_col_types = FALSE)

outDir<-"/Volumes/whipple_lab/Users/cwhilden/PROJECTS/THESIS/chapter2/results/RNAplex"
dir.create(outDir, recursive = TRUE)
makeBedFile(peaks, file.path(outDir,"All_peakCenters.bed"))


```

### Extract peaks 
Use the peak bed coordinates to extract sequences from the reference genome
```{bash, eval=FALSE}
# Interactive session
salloc -p test -c 8 --mem 50G -t 300

# Set working directory
cd /n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2/results/RNAplex

# Activate environment
module load Mambaforge
mamba activate /n/whipple_lab/share_root/Users/cwhilden/ENVS/chimericCW_001

# Generate .fasta files for each peak .bed file
Ref=combined_canonicalTargets.fa

bedtools getfasta -fi $Ref -bed All_peakCenters.bed -s -nameOnly -fo All_peakCenters.fa

mamba deactivate
```


### Snord116 ASE sequences
I selected the 25 nt upstream of the D box (ASE-D) and the D' box (ASE-D') motifs and put them into a fasta file (references/Snord116_Boxes.fa) Now, I will shuffle the ASEs to use as a control

```{r}
# Read fasta file
read_fasta <- function(filepath) {
  lines <- read_lines(filepath)
  tibble(
    name = str_remove(lines[str_detect(lines, "^>")], "^>"),
    seq  = lines[!str_detect(lines, "^>")]
  )
}

fasta<-file.path(macRoot,"references/Snord116_Boxes.fa")
# Generate 5 shuffled versions of each sequence
shuffled_fasta <- read_fasta(fasta) %>%
  rowwise() %>%
  mutate(shuffled = list(replicate(5, paste(sample(strsplit(seq, "")[[1]]), collapse = "")))) %>%
  ungroup() %>%
  unnest_longer(shuffled) %>%
  group_by(name) %>%
  mutate(version = row_number()) %>%
  ungroup() %>%
  mutate(header = paste0(name, "_shuffled_", version))

# Write to new fasta file
write_lines(
  c(rbind(paste0(">", shuffled_fasta$header), shuffled_fasta$shuffled)),
  file.path(macRoot,"results/RNAplex/Snord116_Boxes_Shuffled.fa")
)
```


### Run RNA Plex
Use RNAplex to test snoRNA target basepair complementarity
```{bash, eval=FALSE}
mamba activate /n/whipple_lab/share_root/Users/cwhilden/ENVS/viennaRNA_env

# Canonical target screen
RNAplex -q Snord116_Boxes.fa -t All_peakCenters.fa > RNAplex_results.txt

mamba deactivate

```


### Convert RNA plex output to a csv
Input: RNA plex results in a .txt file
Output: RNA plex results in a readable format (.csv)
```{r}
readRnaPlex <- function(filePath,
                        outputDir,
                        outFileName) {
  x <- readLines(filePath, warn = FALSE)
  
  fastaLines <- grep("^>", x)

  # Get rid of the lines that contain Snord116 (only query used)
  queryLines <- grep("^>Snord116",x)
  
  targetLines <- fastaLines[!fastaLines %in% queryLines]
  
  # each hit is: target line, query line, result line
  # so targetLines should come in steps of 3
  targets <- sub("^>", "", x[targetLines])
  queries <- sub("^>", "", x[queryLines])
  result  <- x[targetLines + 2]

  # structure is everything up to the first occurrence of two+ spaces before coords
  # (structure contains '.' '(' ')' '&')
  structure <- sub("\\s{2,}.*$", "", result)

  # parse "tStart,tEnd  :  qStart,qEnd"
  coordPart <- sub("^.*?\\s{2,}([0-9]+,[0-9]+)\\s+:\\s+([0-9]+,[0-9]+).*$", "\\1|\\2", result)
  tCoord <- sub("\\|.*$", "", coordPart)
  qCoord <- sub("^.*\\|", "", coordPart)

  tStart <- as.integer(sub(",.*$", "", tCoord))
  tEnd   <- as.integer(sub("^.*,", "", tCoord))
  qStart <- as.integer(sub(",.*$", "", qCoord))
  qEnd   <- as.integer(sub("^.*,", "", qCoord))

  # energy inside parentheses, e.g. (-25.30)
  energyStr <- sub("^.*\\((-?[0-9.]+)\\).*$", "\\1", result)
  energy <- as.numeric(energyStr)

  result<-data.frame(
    targetId = targets,
    queryId  = queries,
    structure = structure,
    targetStart = tStart,
    targetEnd   = tEnd,
    queryStart  = qStart,
    queryEnd    = qEnd,
    energy      = energy,
    stringsAsFactors = FALSE
  )
  
  write.csv(result,
            file.path(outputDir,
                      outFileName))
}

# set file paths and output dir
outDir <-outDir

plexFile <-file.path(macRoot,"results/RNAplex/RNAplex_results.txt")


readRnaPlex(plexFile,outDir,"RNAplex_results.csv")

```


## INTERACTION ENERGY V PEAK ENRICHMENT
Read in result files
```{r}
plex <- read_csv(file.path(outDir,"RNAplex_results.csv"))

canonCols <- c('snoRNA','targetId','reads in CLIP', 'reads in INPUT', 'log2_fold_change','negative_log10p','fraction')

peak2 <- peaks %>%
  mutate(targetId=paste0(chromosome,"_",start,"_",end,"()")) %>%
  select(all_of(canonCols))

join <- left_join(peak2, plex,
                           by='targetId',
                           relationship='many-to-many') %>%
  mutate(is_sig=if_else(fraction>=0.1 & log2_fold_change>=1.5,
                        "Sig","Not")) %>%
  mutate(
    # Split structure at & into target and query sides
    target_structure = str_split_fixed(structure, "&", 2)[,1],
    query_structure  = str_split_fixed(structure, "&", 2)[,2],
    
    # Target side metrics
    target_length     = str_length(target_structure),
    target_mismatches = str_count(target_structure, "\\."),
    target_basepairs  = str_count(target_structure, "\\("),
    target_pct_mismatch = (target_mismatches / target_length) * 100,
    
    # Query side metrics
    query_length      = str_length(query_structure),
    query_mismatches  = str_count(query_structure, "\\."),
    query_basepairs   = str_count(query_structure, "\\)"),
    query_pct_mismatch = (query_mismatches / query_length) * 100
  ) %>%
  mutate(snoID=if_else(snoRNA=='Snord116_Snord116l2', 'Snord116', 'other')) %>%
  mutate (category=paste0(is_sig,"_",snoID))

```



```{r}
snord116_dp <- join %>%
  filter(snoRNA=='Snord116_Snord116l2' & queryId=='Snord116_Dpbox')

snord116_dp %>%
ggplot(aes(x=log2(fraction),
           y=-energy)) +
  geom_point() + themeCW()
``` 

