---
title: "Snord116l2_chimericAnalysis"
author: "Courtney Whilden"
date: "2026-03-12"
output: html_document
---

1. Load packages
```{r}
library(tidyverse, quietly=TRUE)
library(ggplot2, quietly=TRUE)
library(tools, quietly=TRUE)
library(ggrepel, quietly=TRUE)
library(ggVennDiagram,quietly = TRUE)
```

2. Set up plot themes
```{r}
# Project directory
clusterRoot <- "/n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2"
macRoot     <- "/Volumes/whipple_lab/Users/cwhilden/PROJECTS/THESIS/chapter2"

# PLOT THEME
paletteCW <- c(
  Pink_CW      = "#EC008C",
  Teal_CW      = "#008C8C",
  Charcoal_CW  = "#2E2E2E",
  LimeGreen_CW = "#9BC53D",
  SoftOrange_CW= "#F4A261",
  Purple_CW    = "#7B2CBF"
)

themeCW <- function(baseSize = 6) {
  theme_minimal(base_size = baseSize) %+replace%
    theme(
      # panel
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(fill = NA, colour = "black", size = 0.5),
      
      # axes lines/ticks/text
      axis.line = element_blank(),                      # border handles the visual frame
      axis.ticks = element_line(size = 0.5),
      axis.ticks.length = unit(0.03, "in"),
      axis.text = element_text(size = baseSize, colour = "black"),
      axis.title = element_text(size = baseSize, colour = "black"),
      
      # legend
      legend.key        = element_blank(),
      legend.background = element_blank(),
      legend.text       = element_text(size = baseSize),
      legend.title = element_blank(),
      
      legend.key.size   = unit(0.2, "cm"),
      legend.spacing.y  = unit(1, "pt"),
      legend.margin     = margin(0, 0, 0, 0, unit = "pt"),
      legend.position = "top",
      
      # plot text & margins
      plot.title = element_text(size = baseSize, colour = "black"),
      plot.subtitle = element_text(size = baseSize, colour = "black"),
      plot.margin = margin(5, 5, 5, 5, unit = "pt"),
      
      aspect.ratio=1
    )
}


######################################
# Function to save plots in my preferred size
plotSave <- function(plotVar, fileName, outSave=outDir){
  ggsave(
  filename = file.path(outSave,fileName),
  plot = plotVar,
  device = "pdf",
  width = 1.5,
  height = 1.5,
  units = "in"
  )
}
```


3. Import Snord116 peaks
Set input paths
```{r}
# iNeuron replicates
data_repA <- read_csv(file.path(macRoot,"summaryFiles/peakSummaryFiles/SSEXPT8_IP_A_S3_R1_001_Significant_Peak_Results.csv"))
data_repB <- read_csv(file.path(macRoot,"summaryFiles/peakSummaryFiles/SSEXPT8_IP_B_S4_R1_001_Significant_Peak_Results.csv"))

# pNeuron probe-enriched
data_probe <- read_csv(file.path(macRoot,"summaryFiles/peakSummaryFiles/TDZ-7_S7_R1_001_Significant_Peak_Results.csv"))

# Read in Nm snoRNA annotation file

Nm <- read_csv(file.path(macRoot,"references/Known_Nm_snoRNAs.csv"))
```

Join iNeuron replicates
```{r}
# Add peak ID containing snoRNA name and peak coordinates
addPeakID <- function(data) {
  
  data$peakID <- paste0(data$snoRNA,";",data$chromosome,"_",data$start,"_",data$end)
  
  return(data)
  
}

data_repA <- addPeakID(data_repA)
data_repB <- addPeakID(data_repB)
data_probe <- addPeakID(data_probe)
# Filter to peakID and fraction values
filterData <- function(data) {
  
  data <- data %>% filter(snoRNA=='Snord116_Snord116l2')
  
  data2 <- data %>%
    select(all_of(c("peakID","fraction","log2_fold_change")))
  
  return(data2) 
  
}

A_filter <- filterData(data_repA)
colnames(A_filter)[2]<-"fracRepA"
colnames(A_filter)[3]<-"l2fcA"

B_filter <- filterData(data_repB)
colnames(B_filter)[2]<-"fracRepB"
colnames(B_filter)[3]<-"l2fcB"

probe_filter <- filterData(data_probe)
colnames(probe_filter)[2]<-"fracProbe"
colnames(probe_filter)[3]<-"l2fcProbe"

# Join replicates into one data frame
data_join <- left_join(A_filter, B_filter, by='peakID')

data_join[is.na(data_join)] <- 0

data_join <- data_join %>%
  mutate(avg_l2fc=(l2fcA+l2fcB)/2,
         avg_frac=(fracRepA+fracRepB)/2)

data_probe <- probe_filter %>%
  mutate(is_sig=if_else(fracProbe>=0.1 & l2fcProbe>=1.5, "Sig","ns"))


data_iN <- data_join %>%
  mutate(is_sig=if_else(avg_frac>=0.1 & avg_l2fc>=1.5, "Sig","ns"))
```

Plot iN replicates
```{r}
plot <- ggplot(data_iN, aes(x = avg_l2fc, y = avg_frac, color = is_sig)) +
  geom_point(alpha = 0.8, size = 1.5) +
  labs(
    x = "log2(Peak fold enrichment)",
    y = "Fraction of chimeras in peak"
  ) +
  geom_text_repel(
    data = subset(data_iN, is_sig=='Sig'),     # Only label significant points
    aes(label = peakID),
    size = 1
  ) +
  geom_vline(xintercept = 1.5) +
  geom_hline(yintercept = 0.1) +
  themeCW()

outDir<-file.path(macRoot,"plots/Snord116_chimeras")
dir.create(outDir, recursive=TRUE)

plotSave(plot,"iNeuron_Snord116Peaks_fraction_v_l2fc.pdf")

plot
```
# Export .bed files of significant peaks:
```{r}
bed_df <- data_iN %>%
  filter(is_sig == "Sig") %>%
  mutate(
    # Get the second part after the semicolon
    peak_coords = sub(".*;", "", peakID),
    # Split into chromosome, start, end
    chrom = sub("_\\d+_\\d+$", "", peak_coords),
    start = as.integer(sub(".*_(\\d+)_\\d+$", "\\1", peak_coords)),
    end = as.integer(sub(".*_(\\d+)$", "\\1", peak_coords))
  ) %>%
  select(chrom, start, end)

# Write to BED file (tab-separated, no header)
write_tsv(bed_df, 
          file.path(macRoot,"summaryFiles/peakSummaryFiles/iNeuron_sigSnord116_peaks.bed"), col_names = FALSE)

bed_df_probe <- data_probe %>%
  filter(is_sig == "Sig") %>%
  mutate(
    # Get the second part after the semicolon
    peak_coords = sub(".*;", "", peakID),
    # Split into chromosome, start, end
    chrom = sub("_\\d+_\\d+$", "", peak_coords),
    start = as.integer(sub(".*_(\\d+)_\\d+$", "\\1", peak_coords)),
    end = as.integer(sub(".*_(\\d+)$", "\\1", peak_coords))
  ) %>%
  select(chrom, start, end)

write_tsv(bed_df_probe, 
          file.path(macRoot,"summaryFiles/peakSummaryFiles/ProbeEnriched_sigSnord116_peaks.bed"), col_names = FALSE)
```


## Export chimeric RPM bw tracks for snord116 chimeras
```{bash, eval=FALSE}
# Start interactive session
salloc -p whipple -c 4 --mem 50G -t 300

# Load environment
module load Mambaforge
mamba activate /n/whipple_lab/share_root/Users/cwhilden/ENVS/chimericCW_001 

cd /n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2
# Run chimeric bam2bw script on desired bam file:
totalChimerasFile="summaryFiles/chimericCounts/ChimerasPerClass_AllSamples.csv"
bamDir="results/mapChimeras"
bwDir="results/Snord116_CPM_BW"
mkdir -p $bwDir

# make rRNA and snRNA tracks
samples=(SSEXPT8_IP_A_S3_R1_001 SSEXPT8_IP_B_S4_R1_001 TDZ-7_S7_R1_001)


for sample in "${samples[@]}"; do

  # rRNA chimeras
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.snoRNA.rRNA.Snord116_Snord116l2.chimeras.bam \
  ${bwDir}/${sample}.Snord116.rRNA.cpm.bw \
  ${bwDir}/${sample}.Snord116.rRNA.cpm.log

  #snRNA chimeras
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.snoRNA.snRNA.Snord116_Snord116l2.chimeras.bam \
  ${bwDir}/${sample}.Snord116.chimeras.snRNA.cpm.bw \
  ${bwDir}/${sample}.Snord116.chimeras.snRNA.cpm.log
  
  # tRNA
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.snoRNA.tRNA.Snord116_Snord116l2.chimeras.bam \
  ${bwDir}/${sample}.Snord116.tRNA.cpm.bw \
  ${bwDir}/${sample}.Snord116.tRNA.cpm.log
  
  
  
  # scaRNA
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.snoRNA.scaRNA.Snord116_Snord116l2.chimeras.bam \
  ${bwDir}/${sample}.Snord116.scaRNA.cpm.bw \
  ${bwDir}/${sample}.Snord116.scaRNA.cpm.log

done

# I also want to grab all snoRNA chimeric files for these samples:
for sample in "${samples[@]}"; do

  # rRNA chimeras
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.snoRNA.rRNA.chimeras.bam \
  ${bwDir}/${sample}.snoRNA.AllChimera.rRNA.cpm.bw \
  ${bwDir}/${sample}.snoRNA.AllChimera.rRNA.cpm.log

  #snRNA chimeras
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.snoRNA.snRNA.chimeras.bam \
  ${bwDir}/${sample}.snoRNA.AllChimera.snRNA.cpm.bw \
  ${bwDir}/${sample}.snoRNA.AllChimera.snRNA.cpm.log
  
  # tRNA
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.snoRNA.tRNA.chimeras.bam \
  ${bwDir}/${sample}.snoRNA.AllChimera.tRNA.cpm.bw \
  ${bwDir}/${sample}.snoRNA.AllChimera.tRNA.cpm.log
  
  
  
  # scaRNA
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.snoRNA.scaRNA.Snord116_Snord116l2.chimeras.bam \
  ${bwDir}/${sample}.snoRNA.AllChimera.scaRNA.cpm.bw \
  ${bwDir}/${sample}.snoRNA.AllChimera.scaRNA.cpm.log

done


```
## Analysis of genome-mapped chimeras
1. Quantify chimeras per gene
Reads overlapping multiple gene ids are not countes
```{bash, eval=FALSE}
# Start interactive session
salloc -p whipple -c 10 --mem 50G -t 300

# Load environment
module load Mambaforge
mamba activate /n/whipple_lab/share_root/Users/cwhilden/ENVS/chimericCW_001 

# Set paths
cd /n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2
gtf=references/gencode.vM38.primary_assembly.annotation.gtf
inDir=results/mapChimeras
outDir=results/Snord116_countGenomeChimeras
mkdir -p $outDir

samples=(SSEXPT8_IP_A_S3_R1_001 SSEXPT8_IP_B_S4_R1_001 TDZ-7_S7_R1_001)
for sample in ${samples[@]}; do
  featureCounts -T 10 -F GTF -a $gtf \
  -o ${outDir}/${sample}.genome.counts.txt \
  ${inDir}/${sample}.snoRNA.genome.Snord116_Snord116l2.chimeras.bam
done

# Count input
clipSample=(SSEXPT8_IN_A_S1_R1_001 SSEXPT8_IN_B_S2_R1_001 TDZ-3_S3_R1_001)
clipIn=results/mapCLIP

for sample in ${clipSample[@]}; do
  featureCounts -T 10 -F GTF -a $gtf \
  -o ${outDir}/${sample}.genome.clip.counts.txt \
  ${clipIn}/${sample}.clip.genomeAligned.out.sort.bam
done


```



## Chimeras per category
```{r}
allChimeras<-read_csv(file.path(macRoot,"summaryFiles/chimericCounts/ChimerasPerClass_AllSamples.csv"))

geneAnnotations<-read_tsv(file.path(macRoot,"references/mus_musculus.gene_info"))
geneAnnotations$Geneid <- sub(".*Ensembl:([^|]+).*", "\\1", geneAnnotations$dbXrefs) 
geneAnnotations<-geneAnnotations%>%
  select(all_of(c('Geneid','Symbol','type_of_gene')))

# Get the number of chimeras that map to protein-coding genes ('pre-mRNA')
breakdownGenome <- function (dataFile,
                             annot=geneAnnotations,
                             columnNames){
  
  data <-read_tsv(dataFile, skip=1) %>%
    select(Geneid,last_col())
  
  colnames(data)<-c('Geneid','counts')

  data_clean <-  data %>%
  mutate(Geneid = sub("\\..*$", "", Geneid))
  
  annot<-left_join(data_clean,annot,by='Geneid') 
  
  annot_summary<- annot %>%
  group_by(type_of_gene) %>%
  summarise(sum(counts))
  
  colnames(annot_summary)<-columnNames
  
  return(annot_summary)
  
}

probe<- breakdownGenome(file.path(macRoot,"results/Snord116_countGenomeChimeras/TDZ-7_S7_R1_001.genome.counts.txt"),
                        columnNames=c('type','TDZ-7_S7_R1_001'))

repA<- breakdownGenome(file.path(macRoot,"results/Snord116_countGenomeChimeras/SSEXPT8_IP_A_S3_R1_001.genome.counts.txt"),
                        columnNames=c('type','SSEXPT8_IP_A_S3_R1_001'))
repB<- breakdownGenome(file.path(macRoot,"results/Snord116_countGenomeChimeras/SSEXPT8_IP_B_S4_R1_001.genome.counts.txt"),
                        columnNames=c('type','SSEXPT8_IP_B_S4_R1_001'))

allGenome <- left_join(probe,repA, by='type') %>%
  left_join(repB,by='type') %>%
  mutate(target=case_when(type=='protein-coding' ~ 'premRNA',
                          type=='pseudo' ~ 'psuedogene',
                          TRUE ~ 'other_counted'))  %>%
  group_by(target) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE))


# pcg
allChimeras <- allChimeras %>%
  select(all_of(c('target','TDZ-7_S7_R1_001','SSEXPT8_IP_A_S3_R1_001','SSEXPT8_IP_B_S4_R1_001'))) %>%
  filter(target !='total')

# Get counts per category: 
# Step 1: Sum allGenome columns to get genome_counted
genome_counted <- allGenome %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE)) %>%
  mutate(target = 'genome_counted') %>%
  relocate(target)

# Step 2: Calculate genome_uncounted (genome from allChimeras - genome_counted)
genome_uncounted <- allChimeras %>%
  filter(target == 'genome') %>%
  pivot_longer(-target, names_to = 'sample', values_to = 'total') %>%
  left_join(
    genome_counted %>% pivot_longer(-target, names_to = 'sample', values_to = 'counted'),
    by = 'sample'
  ) %>%
  mutate(value = total - counted,
         target = 'genome_uncounted') %>%
  select(target, sample, value) %>%
  pivot_wider(names_from = sample, values_from = value)

# Step 3: Combine everything
allCombined <- allChimeras %>%
  filter(target != 'genome') %>%   # remove original genome row
  bind_rows(
            genome_uncounted,
            allGenome)
write.csv(allCombined,file.path(outDir,"Snord116_ChimerasRawPerClass.csv"))
# Convert to long format, calculate CPM, then plot
allCombined_CPM <- allCombined %>%
  pivot_longer(-target, names_to = 'sample', values_to = 'count') %>%
  group_by(sample) %>%
  mutate(CPM = count / sum(count) * 1e6) %>%
  ungroup() %>%
  mutate(target = factor(target, levels = rev(c('rRNA', 'tRNA', 'snRNA', 
                                                'scaRNA', 'premRNA', 'psuedogene',
                                                'genome_counted','other_counted', 'genome_uncounted'))))

wide_percent<- allCombined %>%
  pivot_longer(-target, names_to = 'sample', values_to = 'count') %>%
  group_by(sample) %>%
  mutate(CPM = count / sum(count)) %>%
  ungroup() %>%
  select(target, sample, CPM) %>%
  tidyr::pivot_wider(
    names_from = sample, 
    values_from = CPM
  )

write.csv(wide_percent, file.path(outDir,"Snord116_ChimerasPercentPerClass.csv"))
  
# Stacked bar plot
plot<-ggplot(allCombined_CPM, aes(x = sample, y = CPM, fill = target)) +
  geom_bar(stat = 'identity') +
  labs(x = 'Sample', y = 'Chimeras per Million', fill = 'Target') +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  themeCW()

outDir<-file.path(macRoot,"plots/Snord116_chimeras")
plotSave(plot, "Chimeras_per_targetClass_Snord116.pdf")

plot
```



## Peak calling - genomic chimeras
1. Run clipper
```{bash, eval=FALSE}
out=results/genomePeakCalling
bamDir=results/mapChimeras

sbatch scripts/peakCalling_genomicChimeras.sh $bamDir/SSEXPT8_IP_A_S3_R1_001.snoRNA.genome.Snord116_Snord116l2.chimeras.bam $out
sbatch scripts/peakCalling_genomicChimeras.sh $bamDir/SSEXPT8_IP_B_S4_R1_001.snoRNA.genome.Snord116_Snord116l2.chimeras.bam $out

sbatch scripts/peakCalling_genomicChimeras.sh $bamDir/Fbl_IP_S2_L12_R1_001.snoRNA.genome.Snord116_Snord116l2.chimeras.bam $out
sbatch scripts/peakCalling_genomicChimeras.sh $bamDir/TDZ-7_S7_R1_001.snoRNA.genome.Snord116_Snord116l2.chimeras.bam $out

```
2. Merge peaks


```{bash, eval=FALSE}
salloc -p test -c 4 --mem 50G -t 300

# Load module
module load Mambaforge
mamba activate /net/rcstorenfs/ifs/rc_labs/whipple_lab/share_root/Users/cwhilden/ENVS/chimericCW_001

# Merge biological replicates
cd /n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2
peak=results/genomePeakCalling

fileA=$peak/SSEXPT8_IP_A_S3_R1_001.Snord116.genomePeaks.bed
fileB=$peak/SSEXPT8_IP_B_S4_R1_001.Snord116.genomePeaks.bed

# Output "intersection" file
bedtools intersect -a "$fileA" -b "$fileB" \
  | bedtools sort -i - \
  | bedtools merge -d 1 -c 4,5,6,7,8 -o distinct,min,distinct,min,max \
   > "${peak}/iNeuron_rep_coord_intersect.bed"

# Merge adjacent peak coordinates
bedtools sort -i $peak/Fbl_IP_S2_L12_R1_001.Snord116.genomePeaks.bed | bedtools merge -d 1 -c 4,5,6,7,8 -o distinct,min,distinct,min,max > $peak/Fbl_IP_S2_L12_R1_001.merge.bed
bedtools sort -i $peak/TDZ-7_S7_R1_001.Snord116.genomePeaks.bed| bedtools merge -d 1 -c 4,5,6,7,8 -o distinct,min,distinct,min,max > $peak/TDZ-7_S7_R1_001.merge.bed


mamba deactivate

```

### Annotate peaks with gene info:

mamba activate clipper3_mm39

#using homer
gtf=/n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2/references/gencode.vM38.primary_assembly.annotation.gtf

cd results/genomePeakCalling
annotatePeaks.pl iNeuron_rep_coord_intersect.bed none -gtf $gtf > "iNeuron_rep_coord.annotated.bed"


annotatePeaks.pl Fbl_IP_S2_L12_R1_001.merge.bed none -gtf $gtf > "Fbl_IP_S2_L12_R1_001.merge.annotated.bed"
annotatePeaks.pl TDZ-7_S7_R1_001.merge.bed none -gtf $gtf > "TDZ-7_S7_R1_001.merge.annotated.bed"



### Normalize peaks to input

```{bash, eval=FALSE}

bamDir=results/mapChimeras
inputDir=results/mapCLIP
peakDir=results/genomePeakCalling

sbatch scripts/eCLIP_peakNormalization.sh $bamDir/SSEXPT8_IP_A_S3_R1_001.snoRNA.genome.Snord116_Snord116l2.chimeras.bam \
$inputDir/SSEXPT8_IN_A_S1_R1_001.clip.genomeAligned.out.sort.bam \
$peakDir/iNeuron_rep_coord_intersect.bed \
results/genomePeakCalling

sbatch scripts/eCLIP_peakNormalization.sh $bamDir/SSEXPT8_IP_B_S4_R1_001.snoRNA.genome.Snord116_Snord116l2.chimeras.bam \
$inputDir/SSEXPT8_IN_B_S2_R1_001.clip.genomeAligned.out.sort.bam \
$peakDir/iNeuron_rep_coord_intersect.bed \
results/genomePeakCalling

sbatch scripts/eCLIP_peakNormalization.sh $bamDir/Fbl_IP_S2_L12_R1_001.snoRNA.genome.Snord116_Snord116l2.chimeras.bam \
$inputDir/Fbl_IN_S1_L12_R1_001.clip.genomeAligned.out.sort.bam \
$peakDir/Fbl_IP_S2_L12_R1_001.merge.bed \
results/genomePeakCalling

sbatch scripts/eCLIP_peakNormalization.sh $bamDir/TDZ-7_S7_R1_001.snoRNA.genome.Snord116_Snord116l2.chimeras.bam \
$inputDir/TDZ-3_S3_R1_001.clip.genomeAligned.out.sort.bam \
$peakDir/TDZ-7_S7_R1_001.merge.bed \
results/genomePeakCalling


```



# Summarize peak results
```{r}
annotatePeaks <- function(peakDataFile,
                          peakAnnotFile,
                          outFile){

  peakData<-read_tsv(peakDataFile,col_names=FALSE)
  peakAnnot<-read_tsv(peakAnnotFile)

  colnames(peakData)<-c(
  "chromosome",
  "start",
  "end",
  "name",
  "reads in CLIP",
  "reads in INPUT",
  "p-value",
  "chi_val_or_Fisher",
  "Fisher_orChi_square_test",
  "enriched_or_depleted",
  "negative_log10p",
  "log2_fold_change",
  "entropy")
  
  
  peakClean <- peakData %>%
    mutate(peakID=paste(chromosome,start,end,sep="_"))
  
  annotClean <- peakAnnot %>%
    mutate(peakID=paste(Chr,Start-1,End,sep="_")) %>%
    select(all_of(c('peakID','Gene Name', 'Gene Type','Detailed Annotation')))
  
  annotPeaks <- left_join(peakClean,annotClean,by='peakID')
  
  write.csv(annotPeaks,outFile)
  return(annotPeaks)
}

normPeakDir<-file.path(macRoot,"results/genomePeakCalling/peaksNorm")
annotPeakDir<-file.path(macRoot,"results/genomePeakCalling")
outDir<-file.path(macRoot,"summaryFiles/peakSummaryFiles")
dir.create(outDir,recursive=TRUE)

peakA<-annotatePeaks(file.path(normPeakDir,"SSEXPT8_IP_A_S3_R1_001.snoRNA.genome.Snord116_Snord116l2.chimeras.full"),file.path(annotPeakDir,"iNeuron_rep_coord.annotated.bed"), file.path(outDir,"iNeuron_A_Snord116genomePeaks.csv"))

peakB<-annotatePeaks(file.path(normPeakDir,"SSEXPT8_IP_B_S4_R1_001.snoRNA.genome.Snord116_Snord116l2.chimeras.full"),file.path(annotPeakDir,"iNeuron_rep_coord.annotated.bed"), file.path(outDir,"iNeuron_B_Snord116genomePeaks.csv"))

probe<-annotatePeaks(file.path(normPeakDir,"TDZ-7_S7_R1_001.snoRNA.genome.Snord116_Snord116l2.chimeras.full"),file.path(annotPeakDir,"TDZ-7_S7_R1_001.merge.annotated.bed"), file.path(outDir,"Probe_Snord116genomePeaks.csv"))

```



# Plots

```{r}
data <- probe
plotOut<-"/Volumes/whipple_lab/Users/cwhilden/PROJECTS/THESIS/chapter2/plots/Snord116_chimeras"

exportVolcano<-function(data,
                        outputDir=plotOut,
                        fileName){
  
  plot<-ggplot(data,
       aes(x = log2_fold_change,
           y = negative_log10p,
           color = `Gene Type`)) +
  geom_point(size = 1.5) +
  geom_text_repel(
    data = subset(data, negative_log10p > 30 | log2_fold_change > 10),
    aes(label = `Gene Name`),   # replace gene_name with your label column
    size          = 1,
    max.overlaps  = Inf,
    box.padding   = 0.4,
    point.padding = 0.3,
    segment.color = "grey50",
    show.legend   = FALSE
  ) +
  themeCW()
  
  plotSave(plot,fileName,outputDir)
}
exportVolcano(probe, fileName="ProbeEnriched_genomicPeakSummary.pdf")
exportVolcano(peakA, fileName="RepA_genomicPeakSummary.pdf")
exportVolcano(peakB, fileName="RepB_genomicPeakSummary.pdf")
```


```{r}
getEnriched<-function(data,
                 sampleName){
  sig <- data  %>%
  filter(log2_fold_change>=0 & negative_log10p>=1) %>%
  select(all_of(c(
    "peakID",
    "Gene Name",
    "Gene Type",
    "log2_fold_change",
    "negative_log10p",
    "Detailed Annotation" 
  ))) %>%
    mutate(sample=sampleName)
  
}

probePeaks<-getEnriched(probe,'probeEnriched')
ARepPeaks<-getEnriched(peakA,'RepA')
BRepPeaks<-getEnriched(peakB,'RepB')


allEnriched<-bind_rows(probePeaks,ARepPeaks,BRepPeaks)%>%
  mutate(metaCat=if_else(`Gene Type` %in% c(
    'transcribed_processed_pseudogene',
    'TEC',
    'snRNA',
    'scaRNA',
    'rRNA',
    'Mt_tRNA',
    'Mt_rRNA',
    'misc_RNA',
    'miRNA',
    'IG_V_pseudogene',
    'processed_pseudogene'),
    'Other',
    `Gene Type`))

# Generate a summary by gene type
enriched_summary <- allEnriched %>%
  group_by(sample, metaCat) %>%
  summarise(count = n(), .groups = "drop") %>%
  group_by(sample) %>%
  mutate(proportion = count / sum(count),
         percent    = round(proportion * 100, 1)) %>%
  ungroup() %>%
  mutate(
    sample   = factor(sample,   levels = c("RepA", "RepB", "probeEnriched")),
    metaCat  = factor(metaCat,  levels = c("snoRNA", "protein_coding", 
                                            "lncRNA", "Other"))
  )


```


```{r}
bar<-ggplot(enriched_summary,
       aes(x    = sample,
           y    = count,
           fill = metaCat)) +
  geom_bar(stat     = "identity",
           position = "dodge") +
  geom_text(
    aes(label = count),
    position = position_dodge(width = 0.9),
    vjust    = -0.5,
    size     = 1
  ) +
  labs(
    x     = "Sample",
    y     = "Number of enriched peaks",
    fill  = "Gene Type"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  themeCW()

plotSave(bar,"EnrichedGenomicPeaks_AllSamples.pdf",outSave=plotOut)
```



















## Set significance threshold
```{r}
data<-probe
sig <- data

getSig<-function(data,
                 sampleName){
  sig <- data  %>%
  filter(log2_fold_change>=1.5 & negative_log10p>=10) %>%
  select(all_of(c(
    "peakID",
    "Gene Name",
    "Gene Type",
    "log2_fold_change",
    "negative_log10p",
    "Detailed Annotation" 
  ))) %>%
    mutate(sample=sampleName)
  
  
  
}


probeSig<-getSig(probe,'probeEnriched')
ARepSig<-getSig(peakA,'RepA')
BRepSig<-getSig(peakB,'RepB')


allSig<-bind_rows(probeSig,ARepSig,BRepSig)




ggplot(sig, aes(x = `Gene Type`, fill = `Gene Type`)) +
  geom_bar() +
  geom_text(
    stat  = "count",
    aes(label = after_stat(count)),
    vjust = -0.5,
    size  = 3.5,
    color = "black"
  ) +
  labs(
    x     = "Gene Type",
    y     = "Count",
    title = "Number of Genes by Type"
  ) +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    legend.position = "none"          # color already shown on x-axis
  ) +
  themeCW()


```


```{r}
venn_list <- list(
  "Probe Enriched" = probeSig$`Gene Name`,
  "Rep A"          = ARepSig$`Gene Name`,
  "Rep B"          = BRepSig$`Gene Name`
)
# all three overlap
all_three <- Reduce(intersect, venn_list)

ggVennDiagram(venn_list) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Overlap of Significant Genes") +
  theme(legend.position = "none")
```










