+---
title: "ASO_chimeclip_analysis"
author: "Courtney Whilden"
date: "2026-02-25"
output: html_document
---
The goal of this analysis is to count per-snoRNA chimeras within Nm-snoRNA peaks,
Snord116 peaks, and Snord101 peaks. This will allow me to assess how steric blocking
ASOs affect individual snoRNA interactions.

## Pre-processing
1. Chimeric mapping was done in parallel with other samples:
  chimeric_mapping_peak_calling.Rmd
2. eCLIP done in parallel with other samples:

```{r}
library(tidyverse, quietly=TRUE)
library(ggplot2, quietly=TRUE)
library(tools, quietly=TRUE)
library(ggrepel, quietly=TRUE)
```

```{r setup}
# Project directory
clusterRoot <- "/n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2"
macRoot     <- "/Volumes/whipple_lab/Users/cwhilden/PROJECTS/THESIS/chapter2"

# Import whilden plot settings
plotSave <- function(plotVar, fileName, outSave=outDir, w=1.5, h=1.5){
  ggsave(
  filename = file.path(outSave,fileName),
  plot = plotVar,
  device = "pdf",
  width = w,
  height = h,
  units = "in"
  )
}

# PLOT THEME
paletteCW <- c(
  Pink_CW      = "#EC008C",
  Teal_CW      = "#008C8C",
  Charcoal_CW  = "#2E2E2E",
  LimeGreen_CW = "#9BC53D",
  SoftOrange_CW= "#F4A261",
  Purple_CW    = "#7B2CBF"
)

themeCW <- function(baseSize = 6) {
  theme_minimal(base_size = baseSize) %+replace%
    theme(
      # panel
      panel.background = element_blank(),
      panel.grid.major = element_blank(),
      panel.grid.minor = element_blank(),
      panel.border = element_rect(fill = NA, colour = "black", size = 0.5),
      
      # axes lines/ticks/text
      axis.line = element_blank(),                      # border handles the visual frame
      axis.ticks = element_line(size = 0.5),
      axis.ticks.length = unit(0.03, "in"),
      axis.text = element_text(size = baseSize, colour = "black"),
      axis.title = element_text(size = baseSize, colour = "black"),
      
      # legend
      legend.key        = element_blank(),
      legend.background = element_blank(),
      legend.text       = element_text(size = baseSize),
      legend.title = element_blank(),
      
      legend.key.size   = unit(0.2, "cm"),
      legend.spacing.y  = unit(1, "pt"),
      legend.margin     = margin(0, 0, 0, 0, unit = "pt"),
      legend.position = "top",
      
      # plot text & margins
      plot.title = element_text(size = baseSize, colour = "black"),
      plot.subtitle = element_text(size = baseSize, colour = "black"),
      plot.margin = margin(5, 5, 5, 5, unit = "pt"),
      
      aspect.ratio=1
    )
  
}
```
## Counting

1. I need a .saf file of all significant peaks for the snoRNAs of interest.
Required input: summaryFiles/All_Samples_Significant_Peak_Results.csv

```{r}
clusterRoot<-"/n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2"
macRoot<-"/Volumes/whipple_lab-1/Users/cwhilden/PROJECTS/THESIS/chapter2"


peaks<-read_csv(file.path(macRoot,"summaryFiles/peakSummaryFiles/All_Samples_Significant_Peak_Results.csv"))

# Pull known, Snord101, and Snord116 peaks
peaksFiltered <- peaks %>%
  # Filter for SSEXPT8 samples
  filter(sample=='SSEXPT8_IP_A_S3_R1_001') %>%
  # Keep known, Snord116, and Snord101 peaks
  filter(snoRNA=='Snord101_Gm23130' | snoRNA=='Snord116_Snord116l2' | peakStatus =='known') %>%
  # Keep significant peaks only
  filter(log2_fold_change>=1.5 & fraction>=0.1) %>%
  # Keep rRNA peaks only
  filter(targetClass=='rRNA') %>%
  # Generate peak id
  mutate(peakId=paste0(snoRNA,';',chromosome,'_',start,'_',end))
  

#Build saf file from filtered peak data
saf <- data.frame(
  GeneID=peaksFiltered$peakId,
  Chr=peaksFiltered$chromosome,
  Start=peaksFiltered$start,
  End=peaksFiltered$end,
  Strand='+'
)

# Export saf file
write.table(
  saf,
  file = file.path(macRoot,"summaryFiles/stericblocking/peaks.saf"),
  sep = "\t",
  quote = FALSE,
  row.names = FALSE,
  col.names = TRUE
)

```

2. Count chimeric reads within each peak
```{bash, eval=FALSE}
# Start interactive session
salloc -p whipple -c 10 --mem 50G -t 300

cd /n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2

#Load conda environment
module load Mambaforge
mamba activate /n/whipple_lab/share_root/Users/cwhilden/ENVS/chimericCW_001

# Define data directories
saf=summaryFiles/stericblocking/peaks.saf # saf file of significant peaks
bamDir="results/mapChimeras" # chimeric .bam file directory
bamFiles="${bamDir}"/IGS_CW*.rRNA*.bam # select rRNA chimeric bams from steric block expt
outDir=results/stericBlocking # write output
mkdir -p $outDir

# Run feature counts on all .bam files
for file in $bamFiles; do
  base=$(basename "$file" .bam)

  featureCounts \
    -O \
    -T 8 \
    -F SAF \
    -a $saf \
    -o "${outDir}/${base}.count.txt" \
    "$file"
done

mamba deactivate
exit
```

3. Generate summary file
Required inputs: feature counts files from above code chunk,
summaryFiles/chimericCounts/All_ChimeraCounts.csv
```{r, eval=FALSE}
# Define paths
allChimeraFile <- file.path(macRoot,"summaryFiles/chimericCounts/All_ChimeraCounts.csv")
sampleInfoFile<-file.path(macRoot,"references/sampleAlias.csv")
countDir<-file.path(macRoot,"results/stericBlocking") # output of feature counts

samples <- c("IGS_CW_2IP_CKDL250021054-1A_22W7L3LT4_L4_1.sno",
             "IGS_CW_4IP_CKDL250021054-1A_22W7L3LT4_L4_1.sno",
             "IGS_CW_1IP_CKDL250021054-1A_22W7L3LT4_L4_1.sno",
             "IGS_CW_3IP_CKDL250021054-1A_22W7L3LT4_L4_1.sno"
)


###########
# Function to combine counts across multiple samples into one data frame
# Read in feature counts data
dataFiles <- list.files(path = countDir, 
                        pattern = "\\.count.txt$", 
                        full.names = TRUE)


# Extract sample/snoRNA information from filename
names<-sub("\\.count.txt$","",basename(dataFiles))

# Strip sample and snoRNA from file name
sampleNames <-str_extract(names, "^[^.]+")
snoNames <- str_extract(names, "[^.]+(?=\\.chimeras)")
bamId<-paste0(sampleNames,";",snoNames)

# generate a named list of counts
data <- purrr::set_names(dataFiles, bamId) %>%
  purrr::map(~ readr::read_tsv(.x, show_col_types = FALSE, skip=1))
  
# Merge named list into a single data frame
allData <- data[[1]][,1, drop=FALSE]

for (i in seq_along(bamId)) {
  # Read in counts column (column 7)
  allData[[bamId[i]]] <- data[[i]][,7, drop=TRUE]
  
} 

data_long= allData %>%
  pivot_longer(all_of(bamId),
               names_to="bamId",
               values_to="count_chimera") 
# Read in sample alias
sampleAlias<-read_csv(sampleInfoFile)
# Pull out relevant bam files & chimeras
data_long2 <- data_long %>%
  # Pull peak snoRNA and chimeric bam snoRNA from file names
  separate(Geneid, into = c("peak_snoRNA", "peak_ID"), sep = ";") %>%
  separate(bamId, into = c("sample", "bam_snoRNA"), sep = ";") %>%
  # Match sample alias to sample name
  left_join(sampleAlias, by='sample') %>%
  # Keep rows where the peakID snoRNA equals the bam file snoRNA
  filter(peak_snoRNA==bam_snoRNA) %>%
  # keep relevant count rows
  select(all_of(c(
    'alias',
    'peak_snoRNA',
    'peak_ID',
    'count_chimera'
  )))


#### CPM
# Get total rRNA chimeras for normalization
allChimera<-read_csv(allChimeraFile) %>%
  filter(target=='rRNA') %>%
  filter(snoRNA %in% data_long2$peak_snoRNA) %>%
  select(all_of(samples)) %>%
  # Get rid of .sno from each file name
  rename_with(~ str_remove(.x, "\\.sno"))



# Get rRNA totals for evaluated snoRNAs
chimeraSums <- as.data.frame(colSums(allChimera))
colnames(chimeraSums)<-"count_total"

# Match sample alias
chimeraSums <- chimeraSums %>%
  rownames_to_column("sample") %>%
  left_join(sampleAlias, by = "sample")


######## 
# Match total rRNA chimeric counts
data_long3 <- data_long2 %>%
  left_join(chimeraSums,by='alias') %>%
  mutate(CPM=(count_chimera / count_total) * 1e6,
         peak=paste0(peak_snoRNA,';',peak_ID))


data_final <- data_long3 %>%
  select(all_of(c('alias','CPM','peak','peak_snoRNA'))) %>%
  pivot_wider(
    names_from  = alias,
    values_from = CPM
  ) %>%
  mutate(snoColor=case_when(peak_snoRNA=='Snord116_Snord116l2' ~ 'Snord116',
                   peak_snoRNA=='Snord101_Gm23130' ~ 'Snord101',
                   TRUE ~ 'other'))


write.csv(data_final,
          file.path(macRoot,"summaryFiles/stericblocking/StericBlockCounts.csv"))
```

## Plots
```{r}
macRoot<-"/Volumes/whipple_lab/Users/cwhilden/PROJECTS/THESIS/chapter2"
data<-read_csv(file.path(macRoot,"summaryFiles/stericblocking/StericBlockCounts.csv")) %>%
  filter(if_all(where(is.numeric), ~ .x > 0))

# Calculate log2fc
data$ase101_ctl<-(log2(data$Steric_ase101_RepA/data$Steric_noASO_RepA))
data$aseD_ctl<-(log2(data$Steric_aseD_RepA/data$Steric_noASO_RepA))
data$aseDp_ctl<-(log2(data$Steric_aseDp_RepA/data$Steric_noASO_RepA))


plot101<- data %>%
  ggplot(aes(x=log2(Steric_noASO_RepA),
             y=log2(Steric_ase101_RepA),
             color=snoColor)) +
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous(limits = c(0, NA)) +
  geom_point(size=1.5) +
  geom_abline(slope=1,intercept=0) +
  geom_text_repel(
    data = . %>% filter(snoColor %in% c("Snord101", "Snord116")),
    aes(label = peak),
    size=2) +
  themeCW()

outDir<-file.path(macRoot,"plots/stericblocking")
plotSave(plot101,"ASESnord101_v_noASO_log2CPM.pdf")
plot101
```

```{r}
plot_aseD<- data %>%
  ggplot(aes(x=log2(Steric_noASO_RepA),
             y=log2(Steric_aseD_RepA),
             color=snoColor)) +
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous(limits = c(0, NA)) +
  geom_point(size=1.5) +
  geom_abline(slope=1,intercept=0) +
  geom_text_repel(
    data = . %>% filter(snoColor %in% c("Snord101", "Snord116")),
    aes(label = peak),
    size=2) +
  themeCW()

outDir<-file.path(macRoot,"plots/stericblocking")
plotSave(plot_aseD,"ASED_v_noASO_log2CPM.pdf")
plot_aseD

```

```{r}
plot_aseDp<- data %>%
  ggplot(aes(x=log2(Steric_noASO_RepA),
             y=log2(Steric_aseDp_RepA),
             color=snoColor)) +
  scale_x_continuous(limits = c(0, NA)) +
  scale_y_continuous(limits = c(0, NA)) +
  geom_point(size=1.5) +
  geom_abline(slope=1,intercept=0) +
  geom_text_repel(
    data = . %>% filter(snoColor %in% c("Snord101", "Snord116")),
    aes(label = peak),
    size=2) +
  themeCW()

outDir<-file.path(macRoot,"plots/stericblocking")
plotSave(plot_aseDp,"ASEDp_v_noASO_log2CPM.pdf")
plot_aseDp
```
# waterfall plots of log2 fold change values
```{r}
waterfall <- function(data, sampleColumn) {

  # get symmetric axis limit
  axis_limit <- max(abs(data[[sampleColumn]]), na.rm = TRUE)
  
  data %>%
    arrange(.data[[sampleColumn]]) %>%
    mutate(
      
      label = ifelse(snoColor != "other",peak, NA),
      peak  = factor(peak, levels = peak)
    ) %>%
    ggplot(aes(x = peak, y = .data[[sampleColumn]], fill = snoColor)) +
    geom_bar(stat = "identity", width = 0.8) +
    geom_hline(yintercept = 0, linewidth = 0.4, color = "black") +
    geom_text(aes(label = label,
                  vjust = ifelse(.data[[sampleColumn]] > 0, -0.4, 1.2)),
              size = 1, na.rm = TRUE) +
    scale_y_continuous(limits = c(-axis_limit, axis_limit)) +
    theme_classic() +
    theme(
      axis.text.x  = element_blank(),
      axis.ticks.x = element_blank(),
      legend.position = "top"
    ) +
    labs(
      x     = "Peaks (ranked by fold change)",
      y     = "log2 (ASO / no ASO)",
      fill  = "snoRNA",
      title = paste0("Ranked Fold Change: ", sampleColumn)
    )
}




plotSave(waterfall(data, "ase101_ctl") + themeCW(),
         "ASE101_log2FC_waterfall.pdf",
         w=4, h=2)

plotSave(waterfall(data, "aseD_ctl") + themeCW(),
         "ASED_log2FC_waterfall.pdf",
         w=4, h=2)

plotSave(waterfall(data, "aseDp_ctl") + themeCW(),
         "ASEDp_log2FC_waterfall.pdf",
         w=4, h=2)
```

# Get chimeric RPM bigWigs for IGV

```{bash, eval=FALSE}
# Start interactive session
salloc -p whipple -c 4 --mem 50G -t 300
cd /n/whipple_lab/share_root/Users/cwhilden/PROJECTS/THESIS/chapter2
# Load environment
module load Mambaforge
mamba activate /n/whipple_lab/share_root/Users/cwhilden/ENVS/chimericCW_001 

# Run chimeric bam2bw script on desired bam file:
totalChimerasFile="summaryFiles/chimericCounts/ChimerasPerClass_AllSamples.csv"
bamDir="results/mapChimeras"
bwDir="results/stericBlockingCPM_BW"
mkdir -p $bwDir

# make rRNA and snRNA tracks
samples=(IGS_CW_1IP_CKDL250021054-1A_22W7L3LT4_L4_1 \
IGS_CW_2IP_CKDL250021054-1A_22W7L3LT4_L4_1 \
IGS_CW_3IP_CKDL250021054-1A_22W7L3LT4_L4_1 \
IGS_CW_4IP_CKDL250021054-1A_22W7L3LT4_L4_1
)

for sample in "${samples[@]}"; do

  # snord116
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.snoRNA.rRNA.Snord116_Snord116l2.chimeras.bam \
  ${bwDir}/${sample}.Snord116.rRNA.cpm.bw \
  ${bwDir}/${sample}.Snord116.rRNA.cpm.bw.log

  #snRNA chimeras
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.snoRNA.rRNA.Snord101_Gm23130.chimeras.bam \
  ${bwDir}/${sample}.Snord101.rRNA.cpm.bw \
  ${bwDir}/${sample}.Snord101.rRNA.cpm.bw.log
  
  #all snoRNA chimeras
  scripts/chimericBam2bw_cpm.sh $totalChimerasFile \
  ${sample} \
  ${bamDir}/${sample}.snoRNA.rRNA.chimeras.bam \
  ${bwDir}/${sample}.AllSnoChimeras.rRNA.cpm.bw \
  ${bwDir}/${sample}.AllSnoChimeras.rRNA.cpm.bw.log


done
```
